---
layout: page
title: Vectorization and Joins
reading: ""
notes: "<a href='extras/hw3_merge_ex.R'>hw3_merge_ex.R</a>"
output: 
    ioslides_presentation:
        widescreen: true
        smaller: false
slides: true
---

```{r, echo=FALSE}
options(width = 110)
```


# Vectorization

## Sampling Example {.smaller}

Lets consider a function to generate samples from a bivariate normal.

<br/>

We want to sample `x` and `y` such that 

$$
\left(
\begin{array}{c}
x\\y
\end{array}
\right) \sim N
\left(
\left(
\begin{array}{c}
\mu_x\\\mu_y
\end{array}
\right)
,
\left(
\begin{array}{cc}
\sigma_x^2 & \rho \sigma_x\sigma_y\\
\rho \sigma_x\sigma_y & \sigma_y^2
\end{array}
\right)
\right)
$$

<br/>
we can accomplish this by sampling `x` and `y` separately from

<br/>

$$
\begin{aligned}
x &\;\sim N(\mu_x, \sigma_x^2) \\
y\,|\,x &\;\sim N\left(\mu_y + \frac{\sigma_y}{\sigma_x}\rho(x-\mu_x), (1-\rho^2)\sigma_y^2\right)
\end{aligned}
$$

respectively.

## Nonvectorized rbinorm

```{r}
rbinorm = function(n, mu=c(0,0), sigma=c(1,1), rho=0)
{
  res = matrix(NA, ncol=2, nrow=n)
  colnames(res) = c("x","y")
  for(i in 1:n)
  {
    x = rnorm(1, mu[1], sigma[1])
    y = rnorm(1, mu[2] + (sigma[2]/sigma[1]) * rho * (x - mu[1]), sqrt(1-rho^2) * sigma[2])
    
    res[i,] = c(x,y)
  }

  return(res)
}
```

## Vectorized rbinorm

```{r}
rbinorm_vec = function(n, mu=c(0,0), sigma=c(1,1), rho=0)
{
  x = rnorm(n, mu[1], sigma[1])
  y = rnorm(n, mu[2] + (sigma[2]/sigma[1]) * rho * (x - mu[1]), sqrt(1-rho^2) * sigma[2])
    
  return(cbind(x,y))
}
```

## Output {.smaller}

```{r, fig.width=8, fig.height=4}
par(mfrow=c(1,2), mar=c(4,4,1,1))
plot(rbinorm(1e5,rho=0.7),pch=16,cex=0.2,col=adjustcolor("black",alpha=0.1))
plot(rbinorm_vec(1e5,rho=0.7),pch=16,cex=0.2,col=adjustcolor("black",alpha=0.1))
```

## Microbenchmark

```{r}
library(microbenchmark)
print(microbenchmark(rbinorm(1e5,rho=0.7),rbinorm_vec(1e5,rho=0.7), times=10))
```

## Distance Vectorization {.smaller}

```{r}
gcd = function(loc1, loc2)
{
  deg2rad = function(deg) return(deg*pi/180)

  y1 = deg2rad( loc1[2] )
  y2 = deg2rad( loc2[2] )
  x1 = deg2rad( loc1[1] )
  x2 = deg2rad( loc2[1] )

  R = 6371 # Earth mean radius in km
  
  d = sin(y1)*sin(y2) + cos(y1)*cos(y2) * cos(x2-x1)
  if(d > 1 | d < -1) d = 0
  else d = acos(d) * R
  
  return(d) # distance in km
}
```

## GCD - Non-vectorized {.smaller}

```{r}
gcds = function(l1, l2=l1)
{
  d = matrix(NA, nrow=nrow(l1), ncol=nrow(l2))
  
  for(i in 1:nrow(l1))
  {
    for(j in 1:nrow(l2))
    {
      d[i,j] = gcd(l1[i,],l2[j,])
    }
  }

  return(d)
}
```

## GCD - Vectorized {.smaller}

```{r}
gcd_vec = function(loc1, loc2)
{
  loc1 = matrix(loc1*pi/180,ncol=2)
  loc2 = matrix(loc2*pi/180,ncol=2)

  R = 6371 # Earth mean radius in km

  d = sin(loc1[,2])*sin(loc2[,2]) + cos(loc1[,2])*cos(loc2[,2]) * cos(loc2[,1]-loc1[,1])  
  s = d > 1 | d < -1
  d[s] = 0
  d[!s] = acos(d[!s]) * R
  
  return(d) # distance in km
}

gcds_vec = function(l1, l2=l1)
{
  return( apply(l2, 1, function(x) gcd_vec(l1,x)) )
}

```

## Performance


```{r}
n = 200
locs = cbind(x = runif(n,-180,180), y = runif(n,-90,90))

microbenchmark(gcds(locs),
               gcds_vec(locs),
               times=10)

sum(abs(gcds(locs)-gcds_vec(locs)))
```

# Joins

## Data

<div class="columns-2">
|  Name |       Email       |
|-------|-------------------|
| Alice | alice@company.com |
| Bob   | bob@company.com   |
| Carol | carol@company.com |
| Dave  | dave@company.com  |
| Eve   | eve@company.com   |


|  Name |  Phone       |
|-------|--------------|
| Bob   | 919 555-1111 |
| Carol | 919 555-2222 |
| Eve   | 919 555-3333 |
| Eve   | 310 555-3333 |
| Frank | 919 555-4444 |

</div>

## Outer Join {.smaller}

Outer join of email with phone by name

|  name |           email   |    phone     |
|-------|-------------------|--------------|
| Alice | alice@company.com | NULL         |
|   Bob |   bob@company.com | 919 555-1111 |
| Carol | carol@company.com | 919 555-2222 |
|  Dave |  dave@company.com | NULL         |
|   Eve |   eve@company.com | 919 555-3333 |
|   Eve |   eve@company.com | 310 555-3333 |
| Frank |              NULL | 919 555-4444 |

## Inner Join

Inner join of email with phone by name

|  Name |            Email  |     Phone    |
|-------|-------------------|--------------|
|   Bob |   bob@company.com | 919 555-1111 |
| Carol | carol@company.com | 919 555-2222 |
|   Eve |   eve@company.com | 919 555-3333 |
|   Eve |   eve@company.com | 310 555-3333 |

## Left Join

Left join of email with phone by name

|  name |           email   |    phone     |
|-------|-------------------|--------------|
| Alice | alice@company.com | NULL         |
|   Bob |   bob@company.com | 919 555-1111 |
| Carol | carol@company.com | 919 555-2222 |
|  Dave |  dave@company.com | NULL         |
|   Eve |   eve@company.com | 919 555-3333 |
|   Eve |   eve@company.com | 310 555-3333 |

## Right Join

Right join of email with phone by name

|  name |           email   |    phone     |
|-------|-------------------|--------------|
|   Bob |   bob@company.com | 919 555-1111 |
| Carol | carol@company.com | 919 555-2222 |
|   Eve |   eve@company.com | 919 555-3333 |
|   Eve |   eve@company.com | 310 555-3333 |
| Frank |              NULL | 919 555-4444 |



## Data - R

```{r echo=FALSE}
library(dplyr)
addr = data.frame(name = c("Alice","Bob","Carol","Dave","Eve"),
                  email= c("alice@company.com", "bob@company.com", "carol@company.com", 
                           "dave@company.com", "eve@company.com"))

phone = data.frame(name = c("Bob","Carol","Eve","Eve","Frank"),
                   phone= c("919 555-1111", "919 555-2222", "919 555-3333", 
                            "310 555-3333", "919 555-4444"))
```


<div class="columns-2">
```{r}
addr
```

```{r}
phone
```
</div>

## Outer Join - R

<div class="columns-2">
Base R:
```{r}
merge(addr, phone, all=TRUE)
```

dplyr:

Not implemented, due in 0.3.1.
<br/><br/><br/><br/><br/><br/><br/><br/>
</div>

## Inner Join - R

<div class="columns-2">
Base R:
```{r}
merge(addr, phone, all=FALSE)
```
<br/>

dplyr:
```{r}
inner_join(addr,phone)
```
</div>

## Left Join - R

<div class="columns-2">
Base R:
```{r}
merge(addr, phone, all.x=TRUE)
```
<br/>

dplyr:
```{r}
left_join(addr,phone)
```
</div>

## Right Join - R

<div class="columns-2">
Base R:
```{r}
merge(addr, phone, all.y=TRUE)
```

dplyr:

Not implemented, due in 0.3.1.
<br/><br/><br/><br/><br/><br/><br/><br/>
</div>

## Semi and Anti Joins - R

<div class="columns-2">
```{r}
semi_join(addr, phone)
anti_join(addr, phone)
```
</div>



# Acknowledgments

## Acknowledgments

Above materials are derived in part from the following sources:

* [UseR 2014 - dplyr Tutorial](https://www.dropbox.com/sh/i8qnluwmuieicxc/AAAgt9tIKoIm7WZKIyK25lh6a)
* [dplyr - Introduction Vignette](http://cran.r-project.org/web/packages/dplyr/vignettes/introduction.html)