---
layout: page
title: Subsetting
reading: "<a href='http://adv-r.had.co.nz/Subsetting.html'>Adv-R Subsetting</a>"
notes: ""
output: 
    ioslides_presentation:
        widescreen: true
slides: true
---

## Subsetting in General

R has several different subsetting operators (`[`, `[[`, and `$`).

The behavior of these operators will depend on the object they are being used with.

In general there are 6 different data types that can be used to subset:


<div class="columns-2">
* Positive integers

* Negative integers

* Logical values

* Empty

* Zero

* Character values (names)
</div>

# Subsetting Vectors

## Vectors - Positive Ints {.smaller}

Returns elements at the given location - note R uses a 1-based indexing scheme.

<div class="columns-2">
```{r}
x = c(1,4,7)
x[c(1,3)]
x[c(1,1)]
x[c(1.9,2.1)]
```
<br/><br/><br/>

```{r}
y = list(1,4,7)
str( y[c(1,3)] )
str( y[c(1,1)] )
str( y[c(1.9,2.1)] )
```
</div>

## Vectors - Negative Ints {.smaller}

Excludes elements at the given location

<div class="columns-2">
```{r, error=TRUE}
x = c(1,4,7)
x[-1]
x[-c(1,3)]
x[c(-1,2)]
```
<br/><br/><br/>

```{r, error=TRUE}
y = list(1,4,7)
str( y[-1] )
str( y[-c(1,3)] )
y[c(-1,2)]
```
</div>

## Vectors - Logical Values {.smaller}

Returns elements that correspond to `TRUE` in the logical vector. Length of the logical vector is expected to be the same of the vector being subsetted.

<div class="columns-2">
```{r}
x = c(1,4,7,12)
x[c(TRUE,TRUE,FALSE,TRUE)]
x[c(TRUE,FALSE)]
x[x %% 2 == 0]
```
<br/><br/><br/>

```{r, error=TRUE}
y = list(1,4,7,12)
str( y[c(TRUE,TRUE,FALSE,TRUE)] )
str( y[c(TRUE,FALSE)] )
str( y[y %% 2 == 0] )
```
</div>

## Vectors - Empty

Returns the original vector.

```{r}
x = c(1,4,7)
x[]
y = list(1,4,7)
str(y[])
```

## Vectors - Zero

Returns an empty vector

```{r}
x = c(1,4,7)
x[0]
y = list(1,4,7)
str(y[0])
```

## Vectors - Character Values {.smaller}

If the vector has names, select elements whose names correspond to the character vector.

<div class="columns-2">
```{r}
x = c(a=1,b=4,c=7)
x["a"]
x[c("b","c")]
```

<br/>

```{r}
y = list(a=1,b=4,c=7)
str(y["a"])
str(y[c("b","c")])
```
</div>

## Vectors - Out of bound subsetting {.smaller}


<div class="columns-2">
```{r}
x = c(1,4,7)
x[4]
x["a"]
```

<br/><br/>

```{r}
y = list(1,4,7)
str(y[4])
str(y["a"])
```
</div>

## Vectors - Missing and NULL

<div class="columns-2">
```{r}
x = c(1,4,7)
x[NA]
x[NULL]
```

<br/><br/>

```{r}
y = list(1,4,7)
str(y[NA])
str(y[NULL])
```
</div>

## Vectors - [ vs. [[ {.smaller}

`[[` subsets like `[` except it only subsets a single value. Note that for lists the returned value may not be a list (more on this later).

```{r, error=TRUE}
x = c(1,4,7)
x[[1]]
y = list(1,4,7)
y[2]
y[[2]]
```


## Vectors - [[ vs. $ {.smaller}

`$` is equivalent to `[[` for character subsetting of *lists*, by default it uses partial matching (`exact=FALSE`).

```{r, error=TRUE}
x = c("abc"=1, "def"=5)
x$abc
y = list("abc"=1, "def"=5)
y$abc
y$d
```

## Logical operators and comparisons
  op        Elementwise                  Comp        Elementwise
---------- ----------------- --------- ------------ ---------------
  x | y     True                         x < y        True                    
  x & y     True                         x > y        True     
   !x       True                         x <= y       True         
  x || y    False                        x >= y       True         
  x && y    False                        x != y       True         
 xor(x,y)   True                         x == y       True         
                                         x %in% y     True (for x)                                   



 
## Exercise 1 {.smaller}

Below are 100 values,

```{r}
x = c(56, 3, 17, 2, 4, 9, 6, 5, 19, 5, 2, 3, 5, 0, 13, 12, 6, 31, 10, 21, 8, 4, 1, 1, 2, 5, 16, 1, 3, 8, 1, 
      3, 4, 8, 5, 2, 8, 6, 18, 40, 10, 20, 1, 27, 2, 11, 14, 5, 7, 0, 3, 0, 7, 0, 8, 10, 10, 12, 8, 82, 
      21, 3, 34, 55, 18, 2, 9, 29, 1, 4, 7, 14, 7, 1, 2, 7, 4, 74, 5, 0, 3, 13, 2, 8, 1, 6, 13, 7, 1, 10, 
      5, 2, 4, 4, 14, 15, 4, 17, 1, 9)
```

write down how you would create a subset to accomplish each of the following:

* Select every third value starting at position 2 in `x`.

* Remove all values with an odd index (e.g. 1, 3, etc.)

* Select only the values that are primes. (You may assume all values are less than 100)

* Remove every 4th value, but only if it is odd.




# Matrices, Data Frames, and Arrays

## Matrices and Arrays {.smaller}

Atomic vectors can be treated as multidimensional (2 or more) objects by adding a `dim` attribute.

```{r}
x = 1:8
dim(x) = c(2,4)
x
matrix(1:8, nrow=2, ncol=4)
```

##

<div class="columns-2">
```{r}
x = 1:8
attr(x,"dim") = c(2,2,2)
x
```

```{r}
x = array(1:8,c(2,2,2))
x
```
</div>

## Naming dimensions {.smaller}

```{r}
x = array(1:8,c(2,2,2))
colnames(x) = LETTERS[1:2]
rownames(x) = LETTERS[3:4]
dimnames(x)[[3]] = LETTERS[5:6]
x
```

##

```{r}
str(x)
```

## Subsetting Matrices {.smaller}

```{r}
x = matrix(1:6, nrow=2, ncol=3, dimnames=list(c("A","B"),c("M","N","O")))
```

<div class="columns-2">
```{r}
x[1,3]
x[1:2, 1:2]
x[1:2,]
x[, 1:2]
x[-1,-3]
x[2,-1]
```
</div>

##  {.smaller}

<div class="columns-2">

```{r, error=TRUE}
x["A","M"]
x["A", c("M","O")]
x["B",]
x[, "C"]
x[1,"M"]
x["B"]
x[1]
x[-1]
```
</div>

## Preserving Subsetting

```{r}
x = matrix(1:6, nrow=2, ncol=3, dimnames=list(c("A","B"),c("M","N","O")))
x[1, , drop=FALSE]
x[, 2, drop=FALSE]
```

## Preserving vs Simplifying Subsets

                 Simplifying               Preserving
-------------- ------------------------- -------------------------------------------------------
Vector           `x[[1]]`                  `x[1]`
List             `x[[1]]`                  `x[1]`
Array            `x[1, ]` <br/> `x[, 1]`   `x[1, , drop = FALSE]` <br/> `x[, 1, drop = FALSE]`
Factor           `x[1:4, drop = TRUE]`     `x[1:4]`
Data frame       `x[, 1]` <br/> `x[[1]]`   `x[, 1, drop = FALSE]` <br/> `x[1]`

## Factor Subsetting

```{r}
(x = factor(c("BS", "MS", "PhD", "MS")))
x[1:2]
x[1:2, drop=TRUE]
```

## Data Frame Subsetting {.smaller}

```{r}
df = data.frame(a = 1:2, b = 3:4)
str(df[1])
str(df[[1]])
str(df[, "a", drop = FALSE])
```
## {.smaller}

```{r}
str(df[, "a"])
str(df["a"])
str(df[c("a","b","a")])
str(df[c(FALSE,TRUE)])
```

# Subsetting and assignment

## Subsetting and assignment {.smaller}

Subsets can also be used with assignment to update specific values within an object.

```{r}
x = c(1, 4, 7)
```

```{r}
x[2] = 2
x
x[x %% 2 != 0] = x[x %% 2 != 0] + 1
x
x[c(1,1)] = c(2,3)
x
```

##

```{r}
x = 1:6
```

<div class="columns-2">
```{r, error=TRUE}
x[c(2,NA)] = 1
x
x[c(TRUE,NA)] = 1
x
x[c(-1,-3)] = 3
x
x[] = 6:1
x
```
</div>

## Deleting list (df) elements

```{r}
df = data.frame(a = 1:2, b = TRUE, c = c("A", "B"))
```

```{r}
df[["b"]] = NULL
str(df)
```

```{r}
df[,"c"] = NULL
str(df)
```

## Subsets of Subsets

```{r}
df = data.frame(a = c(5,1,NA,3))
```

```{r}
df$a[df$a == 5] = 0
df[["a"]][df[["a"]] == 1] = 0
df[1][df[1] == 3] = 0
```

```{r}
str(df)
```

## Exercise 2 {.smaller}

Load the course eval data set using the following command:

```{r, eval=FALSE}
d = read.csv("~cr173/Sta523/data/evals.csv")
```

This data frame contains the following variables (columns):

* `cls_val` - students' average course value rating
* `prof_val` - students' average professor value rating
* `rank` - professor's rank (0 - teaching, 1 - tenure track, 2 - tenured)
* `gender` - professor's gender (0 - male, 1 - female)
* `cls_level` - class level (0 - lower division, 1 - upper division)

<br/>

Some of the values in data frame are missing. They have been coded using the value -999, make sure that they are properly treated as `NA`s.

Use subsetting to replace the values of the categorical variables with the appropriate character strings (do not use factors).



# Acknowledgments
## Acknowledgments

Above materials are derived in part from the following sources:

* Hadley Wickham - [Advanced R](http://adv-r.had.co.nz/)
* [R Language Definition](http://stat.ethz.ch/R-manual/R-devel/doc/manual/R-lang.html)
