---
layout: page
title: Vectorization & Recursion
reading: ""
notes: ""
output:
    ioslides_presentation:
        widescreen: true
        smaller: false
slides: true
---

```{r, echo=FALSE}
options(width = 110)
library(microbenchmark)
```


# Vectorization

## Dot product example

```{r}
dot1 = function(x,y) {
  x %*% y
}

dot2 = function(x,y) {
  sum(x * y)
}

dot3 = function(x,y) {
  sum = 0
  for(i in 1:length(x))
    sum = sum + x[i] * y[i]

  return(sum)
}
```

## Dot product benchmark

```{r}
x = rnorm(1e5)
y = runif(1e5)

microbenchmark(dot1(x,y),
               dot2(x,y),
               dot3(x,y),
               times=100)
```


## Sampling Example {.smaller}

Lets consider a function to generate samples from a bivariate normal.

<br/>

We want to sample `x` and `y` such that

$$
\left(
\begin{array}{c}
x\\y
\end{array}
\right) \sim N
\left(
\left(
\begin{array}{c}
\mu_x\\\mu_y
\end{array}
\right)
,
\left(
\begin{array}{cc}
\sigma_x^2 & \rho \sigma_x\sigma_y\\
\rho \sigma_x\sigma_y & \sigma_y^2
\end{array}
\right)
\right)
$$

<br/>
we can accomplish this by sampling `x` and `y` separately from

<br/>

$$
\begin{aligned}
x &\;\sim N(\mu_x, \sigma_x^2) \\
y\,|\,x &\;\sim N\left(\mu_y + \frac{\sigma_y}{\sigma_x}\rho(x-\mu_x), (1-\rho^2)\sigma_y^2\right)
\end{aligned}
$$

respectively.

## Nonvectorized rbinorm {.smaller}

```{r}
rbinorm = function(n, mu=c(0,0), sigma=c(1,1), rho=0)
{
  res = matrix(NA, ncol=2, nrow=n)
  colnames(res) = c("x","y")
  for(i in 1:n)
  {
    x = rnorm(1, mu[1], sigma[1])
    y = rnorm(1, mu[2] + (sigma[2]/sigma[1]) * rho * (x - mu[1]), sqrt(1-rho^2) * sigma[2])

    res[i,] = c(x,y)
  }

  return(res)
}
```

## Vectorized rbinorm {.smaller}  

```{r}
rbinorm_vec = function(n, mu=c(0,0), sigma=c(1,1), rho=0)
{
  x = rnorm(n, mu[1], sigma[1])
  y = rnorm(n, mu[2] + (sigma[2]/sigma[1]) * rho * (x - mu[1]), sqrt(1-rho^2) * sigma[2])

  return(cbind(x,y))
}
```

## Output {.smaller}

```{r, fig.width=8, fig.height=4}
par(mfrow=c(1,2), mar=c(4,4,1,1))
plot(rbinorm(1e5,rho=0.7),pch=16,cex=0.2,col=adjustcolor("black",alpha=0.1))
plot(rbinorm_vec(1e5,rho=0.7),pch=16,cex=0.2,col=adjustcolor("black",alpha=0.1))
```

## Benchmarking

```{r}
print(microbenchmark(rbinorm(1e5,rho=0.7),rbinorm_vec(1e5,rho=0.7), times=10))
```

## GCD - Non-vectorized {.smaller}

```{r}
gcd = function(loc1, loc2) {
  deg2rad = function(deg) return(deg*pi/180)

  y1 = deg2rad( loc1[2] )
  y2 = deg2rad( loc2[2] )
  x1 = deg2rad( loc1[1] )
  x2 = deg2rad( loc2[1] )

  R = 6371 # Earth mean radius in km

  d = sin(y1)*sin(y2) + cos(y1)*cos(y2) * cos(x2-x1)
  if(d > 1 | d < -1) d = 0
  else d = acos(d) * R

  return(d) # distance in km
}

gcds = function(l1, l2=l1) {
  d = matrix(NA, nrow=nrow(l1), ncol=nrow(l2))

  for(i in 1:nrow(l1)) {
    for(j in 1:nrow(l2)) {
      d[i,j] = gcd(l1[i,],l2[j,])
    }
  }

  return(d)
}
```

## GCD - Vectorized {.smaller}

```{r}
gcd_vec = function(loc1, loc2)
{
  loc1 = matrix(loc1*pi/180,ncol=2)
  loc2 = matrix(loc2*pi/180,ncol=2)

  R = 6371 # Earth mean radius in km

  d = sin(loc1[,2])*sin(loc2[,2]) + cos(loc1[,2])*cos(loc2[,2]) * cos(loc2[,1]-loc1[,1])
  s = d > 1 | d < -1
  d[s] = 0
  d[!s] = acos(d[!s]) * R

  return(d) # distance in km
}

gcds_vec = function(l1, l2=l1)
{
  return( apply(l2, 1, function(x) gcd_vec(l1,x)) )
}

```

## Performance


```{r}
n = 200
locs = cbind(x = runif(n,-180,180), y = runif(n,-90,90))

microbenchmark(gcds(locs),
               gcds_vec(locs),
               times=10)

sum(abs(gcds(locs)-gcds_vec(locs)))
```

# Recursion 

## Live Demo-
