---
layout: page
title: Regular expressions and stringr
reading: "<a href='http://www.cheatography.com/davechild/cheat-sheets/regular-expressions/'>RegEx Cheat Sheet</a> and <a href='https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html'>stringr</a>"
notes: "<a href='http://regexr.com/'>Testing RegExs</a>"
output: 
    ioslides_presentation:
        widescreen: true
slides: true
link: true
---

```{r echo=FALSE}
suppressMessages(library(stringr))
```

# stringr

## stringr {.smaller}

stringr is a string handling package written by Hadley Wickham that is designed to improve / simplify string handling in R. 

|     | |
|-----|-|
|`str_detect(string, pattern)`| Detect the presence or absence of a pattern in a string. |
|`str_locate(string, pattern)`| Locate the first position of a pattern and return a matrix with start and end. |
|`str_extract(string, pattern)`| Extracts text corresponding to the first match. |
|`str_match(string, pattern)`| Extracts capture groups formed by `()` from the first match. |
|`str_split(string, pattern)`| Splits string into pieces and returns a list of character vectors. |
|`str_replace(string, pattern, replacement)`| Replaces the first matched pattern and returns a character vector. |

<br/>

Many of these functions have variants with an `_all` suffix which will match more than one occurrence of the pattern in a given string.


# Regular Expressions

## Simple Matching

```{r}
text = c("The","quick","brown","fox","jumps","over","the","lazy","dog")
```

```{r}
str_detect(text,"quick")
str_detect(text,"o")
str_detect(text,"row")
```

## Metacharacters {.smaller}

The power of regular expressions comes from the ability to use a number of special metacharacters that modify how the pattern matching is performed.

```regex
. ^ $ * + ? { } [ ] \ | ( )
```

Because of their special properties they cannot be matched directly, if you need to match one of them you need to escape it first (precede it by `\`). One problem is that 

```{r error=TRUE}
str_detect("abc[def","[")
str_detect("abc[def","\\[")
```

Note we actually need to prefix `[` with `\\` because `\` is also a escape character for normal strings.


## Anchors

|      | |
|------|-|
| `^` or `\A` | Start of string   |
| `$` or `\Z` | End of string     |
| `\b`        | Word boundary     |  
| `\B`        | Not word boundary |

## Anchor Examples {.smaller}

```{r}
text = "The quick brown fox jumps over the lazy dog"
```
<div class="columns-2">
```{r}
str_replace(text,"^The","-")
str_replace(text,"^dog","-")
str_replace(text,"The$","-")
str_replace(text,"dog$","-")
str_replace(text,"\\AThe","-")
str_replace(text,"dog\\Z","-")
str_replace(text,"\\bbrown\\b","-")
str_replace(text,"\\Brow\\B","-")
```
</div>


## Character Classes

|      | |
|------|-|
| `\s` | White space |
| `\S` | Not white space |
| `\d` | Digit (0-9)|
| `\D` | Not digit |
| `\w` | Word (A-Z, a-z, 0-9, or _) |
| `\W` | Not word |

## Ranges

|          | |
|----------|-|
| `.`      | Any character except new line (\n) |
| `[abc]`  | Range (a or b or c) |
| `[^abc]` | Not (a or b or c) |
| `[a-q]`  | Lower case letter from a to q |
| `[A-Q]`  | Upper case letter from A to Q |
| `[0-7]`  | Digit from 0 to 7 |

## Quantifiers

Attached to literals, character classes, ranges or groups to match repeats.

|         | |
|---------|-|
| `*`     | Match 0 or more |
| `+`     | Match 1 or more |
| `?`     | Match 0 or 1 |
| `{3}`   | Match Exactly 3 |
| `{3,}`  | Match 3 or more |
| `{3,5}` | Match 3, 4 or 5 |

## Greedy vs ungreedy matching 

Add a ? to a quantifier to make it ungreedy.

```{r}
text = "<div class='main'> <div> <a href='here.pdf'>Here!</a> </div> </div>"
```

```{r}
str_extract(text, "<div>.*</div>")
str_extract(text, "<div>.*?</div>")
```

## Groups

Group together parts of a regular expression for modification or capture.

|           | |
|-----------|-|
| `(a|b)`   | match literal a or b, group either
| `a(bc)?`  | match literal a or abc, group bc or ""
| `(?:abc)` | Non-c­apt­uring group
| `(abc)def(hig)| match abcdefhig, group abc and hig

## Group Examples {.smaller}

```{r}
text = c("apple", "219 733 8965", "329-293-8753", "Work: 579-499-7527; Home: 543.355.3679")
phone = "([2-9][0-9]{2})[- .]([0-9]{3})[- .]([0-9]{4})"
```

```{r}
str_extract(text, phone)
str_extract_all(text, phone)
```
## Matching {.smaller}

```{r}
str_match(text, phone)
str_match_all(text, phone)
```

## Exercise {.smaller}

Write one or more regular expressions to extract the data contained in the xml file below:

```xml
<breakfast_menu>
  <food>
    <name>Belgian Waffles</name>
    <price>$5.95</price>
    <calories>650</calories>
    <addon>
        <name>Strawberries</name>
        <price>$2.00</price>
        <calories>250</calories>
    </addon>
  </food>
  <food>
    <name>French Toast</name>
    <price>$4.50</price>
    <calories>600</calories>
    <addon>
        <name>Strawberries</name>
        <price>$2.00</price>
        <calories>250</calories>
    </addon>
  </food>
</breakfast_menu>
```

# Acknowledgments

## Acknowledgments

* Hadley Wickham - [stringr vigneete](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html)

* David Child - [RegEx Cheat Sheet](http://www.cheatography.com/davechild/cheat-sheets/regular-expressions/)

* [Regular-Expression.info](http://www.regular-expressions.info/)